Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-06-12 manuscript.Rmd
Implementing revisions

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. This novel paradigm allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Periodic flicker entrains both neural [@Norcia2015] and pupil [@Spitschan2014] responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. We followed up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e. temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways implement stronger interocular suppression than the canonical cortical pathway.

@@

We initially preregistered our main hypotheses and analysis intentions for the first experiment. We then conducted a pilot study with N=12 participants, before making some minor changes to the stimulus (we added dim red lines to aid binocular fusion). We then ran the main experiment, followed by two additional experiments that were not preregistered. The preregistration document, raw data files, and experimental and analysis code, are available on the project repository: https://doi.org/10.17605/OSF.IO/TBEMA.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [e.g. @Anstis1998; @Levelt1965; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e. sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation at high contrasts [@Baker2017; @Moradi2009; @Meese2006]. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, @Quaia2018 observed a strong binocular facilitation (or 'supersummation') in the reflexive eye movement response to rapidly moving stimuli, and @Spitschan2019 report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm [@Baker2020]. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments [@Levelt1965; @Anstis1998; @Baker2012], though not for luminance decrements [@Anstis1998]. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

@@

Figures \@ref(fig:appendixfig1) and \@ref(fig:appendixfig2)

```{r appendixfig1, fig.cap="S.", fig.align="center", echo=FALSE}

@@

```{r appendixfig2, fig.cap="S.", fig.align="center", echo=FALSE}

>>>>

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. We chose a primary flicker frequency of 2Hz as a compromise between the low-pass pupil response [see @Spitschan2014], and the relatively higher-pass EEG response [@Regan1966]. This novel paradigm allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Periodic flicker entrains both neural [@Norcia2015] and pupil [@Spitschan2014] responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. Relative to the response to a monocular signal, adding a signal in the other eye can either increase the response (facilitation) or reduce it (suppression). We followed up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e. temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways implement stronger interocular suppression than the canonical cortical pathway.

@@

We initially preregistered our main hypotheses and analysis intentions for the first experiment. We then conducted a pilot study with N = 12 participants, before making some minor changes to the stimulus (we added dim red lines to aid binocular fusion). We then ran the main experiment, followed by two additional experiments that were not preregistered. The preregistration document, raw data files, and experimental and analysis code, are available on the project repository: https://doi.org/10.17605/OSF.IO/TBEMA.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [e.g. @Anstis1998; @Levelt1965; @Legge1981; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e. sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation at high contrasts [@Baker2017; @Moradi2009; @Meese2006]. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, @Quaia2018 observed a strong binocular facilitation (or 'supersummation') in the reflexive eye movement response to rapidly moving stimuli, and @Spitschan2019 report a similar result in archival data on melatonin suppression due to light exposure. Work on the accommodative response indicates that binocular combination there is approximately linear [@Flitcroft1992], and can even cancel when signals are in antiphase (we did not try this configuration here). In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm [@Baker2020]. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments [@Levelt1965; @Anstis1998; @Baker2012], though not for luminance decrements [@Anstis1998]. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

@@

We conducted a conceptual replication of Experiment 1 using an alternative system. We used a pair of Spectra Tune Lab multiprimary devices (LEDmotive Technologies LLC, Barcelona, Spain), coupled to a binocular headset using liquid light guides. The light was imaged onto a circular diffuser for each eye (field size X deg) with the central 6 degrees masked off using a black occluder. Therefore the replication experiment involved peripheral stimulation, unlike the main experiment which stimulated the central ~4 degrees of the visual field. The mean luminance of the devices was $cd/m^2$. All conditions were otherwise as described for the main experiment, and we tested 12 participants in total.

The pupillometry results are shown in Figure \@ref(fig:appendixfig1) and correspond closely to those from the main experiment (Figure \@ref(fig:pupildata)). The ratio of binocular to monocular responses in Figure \@ref(fig:appendixfig1)c is similar, and suppression is evident in Figure \@ref(fig:appendixfig1)d,e. Of particular interest is the presence of a slight response at the second harmonic (4Hz, Figure \@ref(fig:appendixfig1)b), which was not present in our original data. This may be because the driving signal is clearer when stimulating the periphery (note the clear waveform in Figure \@ref(fig:appendixfig1)a), or it might indicate additional nonlinearities not present in the fovea.

```{r appendixfig1, fig.cap="Summary of pupillometry results for N = 12 participants, for peripheral stimulation. See Figure 1 for an overview of each panel.", fig.align="center", echo=FALSE}

@@

The EEG results are shown in Figure \@ref(fig:appendixfig2). These still show a strong binocular facilitation effect at the highest contrast levels (Figure \@ref(fig:appendixfig2)c,f), but the contrast response function is less clear at both the first and second harmonics. We suspect that this is because the cortical representation of the peripheral visual field is primarily along the calcarine sulcus, which results in some cancellation of the steady-state signal. This results in weaker signals than we obtained for foveal stimulation (represented at the occipital pole) in the main experiment (see Figure \@ref(fig:EEGdata)).

```{r appendixfig2, fig.cap="Summary of steady-state EEG results for N = 12 participants, for peripheral stimulation. See Figure 2 for an overview of each panel.", fig.align="center", echo=FALSE}

### 2023-06-12 manuscript.tex
Implementing revisions

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. This novel paradigm allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Periodic flicker entrains both neural (\protect\hyperlink{ref-Norcia2015}{Norcia et al., 2015}) and pupil (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}) responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. We followed up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e.~temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways implement stronger interocular suppression than the canonical cortical pathway.

@@

We initially preregistered our main hypotheses and analysis intentions for the first experiment. We then conducted a pilot study with N=12 participants, before making some minor changes to the stimulus (we added dim red lines to aid binocular fusion). We then ran the main experiment, followed by two additional experiments that were not preregistered. The preregistration document, raw data files, and experimental and analysis code, are available on the project repository: \url{https://doi.org/10.17605/OSF.IO/TBEMA}.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. \protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}; \protect\hyperlink{ref-Quaia2018}{Quaia et al., 2018}). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e.~sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation at high contrasts (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}; \protect\hyperlink{ref-Moradi2009}{Moradi and Heeger, 2009}). Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al. (\protect\hyperlink{ref-Quaia2018}{2018}) observed a strong binocular facilitation (or `supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and Cajochen (\protect\hyperlink{ref-Spitschan2019}{2019}) report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm (\protect\hyperlink{ref-Baker2020}{Baker et al., 2020}). Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments (\protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Baker2012}{Baker et al., 2012}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}), though not for luminance decrements (\protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}). Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

@@

@@

@@

@@

Figures \ref{fig:appendixfig1} and \ref{fig:appendixfig2}

@@

\caption{S.}\label{fig:appendixfig1}
\end{figure}

@@

\caption{S.}\label{fig:appendixfig2}
\end{figure}

>>>>

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. We chose a primary flicker frequency of 2Hz as a compromise between the low-pass pupil response (see \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}), and the relatively higher-pass EEG response (\protect\hyperlink{ref-Regan1966}{Regan, 1966}). This novel paradigm allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Periodic flicker entrains both neural (\protect\hyperlink{ref-Norcia2015}{Norcia et al., 2015}) and pupil (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}) responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. Relative to the response to a monocular signal, adding a signal in the other eye can either increase the response (facilitation) or reduce it (suppression). We followed up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e.~temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways implement stronger interocular suppression than the canonical cortical pathway.

@@

We initially preregistered our main hypotheses and analysis intentions for the first experiment. We then conducted a pilot study with N = 12 participants, before making some minor changes to the stimulus (we added dim red lines to aid binocular fusion). We then ran the main experiment, followed by two additional experiments that were not preregistered. The preregistration document, raw data files, and experimental and analysis code, are available on the project repository: \url{https://doi.org/10.17605/OSF.IO/TBEMA}.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. \protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Legge1981}{Legge and Rubin, 1981}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}; \protect\hyperlink{ref-Quaia2018}{Quaia et al., 2018}). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e.~sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation at high contrasts (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}; \protect\hyperlink{ref-Moradi2009}{Moradi and Heeger, 2009}). Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al. (\protect\hyperlink{ref-Quaia2018}{2018}) observed a strong binocular facilitation (or `supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and Cajochen (\protect\hyperlink{ref-Spitschan2019}{2019}) report a similar result in archival data on melatonin suppression due to light exposure. Work on the accommodative response indicates that binocular combination there is approximately linear (\protect\hyperlink{ref-Flitcroft1992}{Flitcroft et al., 1992}), and can even cancel when signals are in antiphase (we did not try this configuration here). In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm (\protect\hyperlink{ref-Baker2020}{Baker et al., 2020}). Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments (\protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Baker2012}{Baker et al., 2012}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}), though not for luminance decrements (\protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}). Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Flitcroft1992}{}}%
Flitcroft DI, Judge SJ, Morley JW. 1992. Binocular interactions in accommodation control: Effects of anisometropic stimuli. \emph{J Neurosci} \textbf{12}:188--203. doi:\href{https://doi.org/10.1523/JNEUROSCI.12-01-00188.1992}{10.1523/JNEUROSCI.12-01-00188.1992}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Legge1981}{}}%
Legge GE, Rubin GS. 1981. Binocular interactions in suprathreshold contrast perception. \emph{Percept Psychophys} \textbf{30}:49--61. doi:\href{https://doi.org/10.3758/bf03206136}{10.3758/bf03206136}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Regan1966}{}}%
Regan D. 1966. Some characteristics of average steady-state and transient responses evoked by modulated light. \emph{Electroencephalogr Clin Neurophysiol} \textbf{20}:238--48. doi:\href{https://doi.org/10.1016/0013-4694(66)90088-5}{10.1016/0013-4694(66)90088-5}

@@

We conducted a conceptual replication of Experiment 1 using an alternative system. We used a pair of Spectra Tune Lab multiprimary devices (LEDmotive Technologies LLC, Barcelona, Spain), coupled to a binocular headset using liquid light guides. The light was imaged onto a circular diffuser for each eye (field size X deg) with the central 6 degrees masked off using a black occluder. Therefore the replication experiment involved peripheral stimulation, unlike the main experiment which stimulated the central \textasciitilde4 degrees of the visual field. The mean luminance of the devices was \(cd/m^2\). All conditions were otherwise as described for the main experiment, and we tested 12 participants in total.

The pupillometry results are shown in Figure \ref{fig:appendixfig1} and correspond closely to those from the main experiment (Figure \ref{fig:pupildata}). The ratio of binocular to monocular responses in Figure \ref{fig:appendixfig1}c is similar, and suppression is evident in Figure \ref{fig:appendixfig1}d,e. Of particular interest is the presence of a slight response at the second harmonic (4Hz, Figure \ref{fig:appendixfig1}b), which was not present in our original data. This may be because the driving signal is clearer when stimulating the periphery (note the clear waveform in Figure \ref{fig:appendixfig1}a), or it might indicate additional nonlinearities not present in the fovea.

@@

\caption{Summary of pupillometry results for N = 12 participants, for peripheral stimulation. See Figure 1 for an overview of each panel.}\label{fig:appendixfig1}
\end{figure}

The EEG results are shown in Figure \ref{fig:appendixfig2}. These still show a strong binocular facilitation effect at the highest contrast levels (Figure \ref{fig:appendixfig2}c,f), but the contrast response function is less clear at both the first and second harmonics. We suspect that this is because the cortical representation of the peripheral visual field is primarily along the calcarine sulcus, which results in some cancellation of the steady-state signal. This results in weaker signals than we obtained for foveal stimulation (represented at the occipital pole) in the main experiment (see Figure \ref{fig:EEGdata}).

@@

\caption{Summary of steady-state EEG results for N = 12 participants, for peripheral stimulation. See Figure 2 for an overview of each panel.}\label{fig:appendixfig2}
\end{figure}
